---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# eq5d

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/eq5d)](https://CRAN.R-project.org/package=eq5d)
[![Downloads](https://cranlogs.r-pkg.org/badges/grand-total/eq5d)](https://CRAN.R-project.org/package=eq5d)
[![R-CMD-check](https://github.com/fragla/eq5d/actions/workflows/check-standard.yaml/badge.svg)](https://github.com/fragla/eq5d/actions/workflows/check-standard.yaml)
[![Codecov test coverage](https://codecov.io/gh/fragla/eq5d/branch/master/graph/badge.svg)](https://app.codecov.io/gh/fragla/eq5d?branch=master)
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](LICENSE.md)
<!-- badges: end -->

`eq5d` provides tools for the valuation, reporting and analysis of EQ-5D 
health outcomes.

The package supports utility index calculation, descriptive system reporting, 
severity and distributional summaries, longitudinal change analysis and 
visualisation of EQ-5D profiles. A companion Shiny application is also included 
for interactive analysis of EQ-5D datasets.

EQ-5D is a health-related quality-of-life instrument developed by the EuroQol 
Group. Health is described using five dimensions: Mobility (MO), 
Self-care (SC), Usual Activities (UA), Pain/Discomfort (PD) and 
Anxiety/Depression (AD). `eq5d` supports valuation and analysis of EQ-5D-3L, 
EQ-5D-Y-3L and EQ-5D-5L data.

## Installation

Install the latest release from [CRAN](https://CRAN.R-project.org/package=eq5d):

```r
install.packages("eq5d")
```

Install the development version from [GitHub](https://github.com/fragla/eq5d):

```r
# install.packages("remotes")
remotes::install_github("fragla/eq5d")
```

## Getting started

The examples below illustrate some of the main functionality provided by `eq5d`.

### Utility index calculation

```{r}
library(eq5d)

scores <- c(MO = 1, SC = 2, UA = 3, PD = 2, AD = 1)

eq5d(scores, country = "UK", version = "3L", type = "TTO")

# Five-digit health state
eq5d(12321, country = "UK", version = "3L", type = "TTO")
```

### Finding value sets

```{r}
valuesets(
  country = "France",
  references = c("PubMed", "DOI")
)
```

### Descriptive-system reporting

```{r}
dat <- read.csv(
  system.file("extdata", "eq5d3l_example.csv", package = "eq5d")
)

dd <- descriptive_data(
  dat,
  version = "3L",
  metric = "percent"
)

table_descriptive(dd)
```

### Severity summaries

```{r}
lss(scores, version = "3L")

lfs(scores, version = "3L")
```

### Change analysis

```{r}
pre  <- subset(dat, Group == "Group1")
post <- subset(dat, Group == "Group2")

pchc(pre, post, version = "3L", summary = TRUE)
```

## Value sets

`eq5d` includes a wide range of EQ-5D-3L, EQ-5D-5L, EQ-5D-Y-3L, crosswalk,
reverse-crosswalk and DSU mapping value sets.

See `?valuesets` for details.

## Documentation

Further documentation is available in the package vignettes:

**Core reporting vignettes**

- Reporting EQ-5D Data with `eq5d`
- Reporting the EQ-5D Descriptive System
- Reporting EQ-5D Severity and Distributional Summaries
- Reporting EQ-5D Change Analysis

**Specialised vignettes**

- Mapping between EQ-5D-5L and EQ-5D-3L using the NICE DSU models

## Shiny application

Launch the companion Shiny application with:

```r
shiny_eq5d()
```

Online version:

https://fragla.shinyapps.io/shiny-eq5d

## Contributing

Bug reports, feature requests and pull requests are welcome.

Please use the GitHub issue tracker:

https://github.com/fragla/eq5d/issues

## Citation

To obtain citation information for `eq5d`:

```r
citation("eq5d")
```

## License

MIT License.
