---
title: "Calculating EQ-5D indices with eq5d"
author: "Fraser Morton and Jagtar Singh Nijjar"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{eq5d}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

#EQ-5D

EQ-5D is a one of the most popular health related quality of life instruments that can be used in the clinical and economic evaluation of health care as well as in population health surveys. Developed by the [EuroQol](https://www.euroqol.org) group, the instrument consists of two components: health state description and evaluation.  

In the description component a person's health status is measured in terms of five dimensions; mobility, self-care, usual activities, pain/discomfort, and anxiety/depression. For each dimension a patient self-rates their level of severity for using a three-level ([EQ-5D-3L](https://euroqol.org/eq-5d-instruments/eq-5d-3l-about/)) or a five-level ([EQ-5D-5L](https://euroqol.org/eq-5d-instruments/eq-5d-5l-about/)) scale, generating a five digit number.

The evaluation component requires a patient to record their overall health status using a visual analogue scale (EQ-VAS).

Following assessment the five digit number from the descriptive component can be converted to an index score to compare how good or bad the patient's health state is relative to the general population of a country.

The package eq5d provides methods to calculate index scores for both EQ-5D-3L and EQ-5D-5L value sets. Additionally, a [Shiny](https://shiny.rstudio.com) app is provided to enable multiple index values to be calculated via a web browser.

## Quick Start

```{r quickStart}
library(eq5d)

scores <- c(Mobility=1,Care=2,Activity=3,Pain=4,Anxiety=5)

eq5d(scores=scores, country="Indonesia", version="5L")
```

```{r valuesets}
# Return all value sets.
head(valuesets())

# Return VAS based value sets.
head(valuesets(type="VAS"))

# Return EQ-5D-5L value sets.
head(valuesets(version="5L"))

# Return all UK value sets.
valuesets(country="UK")
```


```{r overview}
eq5d(scores=c(Mobility=3,Care=2,Activity=3,Pain=2,Anxiety=3), 
  type="TTO", version="3L", country="Germany")

```


