---
title: "Calculating EQ-5D indices and summarising profiles with eq5d"
author: "Fraser Morton and Jagtar Singh Nijjar"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculating EQ-5D indices and summarising profiles with eq5d}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# EQ-5D

EQ-5D is a popular health related quality of life instrument used in the clinical and economic evaluation of health care. Developed by the [EuroQol](https://euroqol.org/) group, the instrument consists of two components: health state description and evaluation. 

For the description component a subject self-rates their health in terms of five dimensions; mobility, self-care, usual activities, pain/discomfort, and anxiety/depression using either a three-level ([EQ-5D-3L](https://euroqol.org/eq-5d-instruments/eq-5d-3l-about/)) or a five-level ([EQ-5D-5L](https://euroqol.org/eq-5d-instruments/eq-5d-5l-about/)) scale. 

The evaluation component requires a patient to record their overall health status using a visual analogue scale (EQ-VAS).

Following assessment the scores from the descriptive component can be reported as a five digit number ranging from 11111 (full health) to 33333/55555 (worst health). A number of methods exist for analysing these five digit profiles. However, frequently they are converted to a single utility index using country specific value sets, which can be used in the clinical and economic evaluation of health care as well as in population health surveys.

The eq5d package provides methods for the cross-sectional and longitudinal analysis of EQ-5D profiles and also the calculation of utility index scores from a subject's dimension scores. Additionally, a [Shiny](https://shiny.rstudio.com) app is included to enable the calculation, visualisation and automated statistical analysis of multiple EQ-5D index values via a web browser using EQ-5D dimension scores stored in CSV or Excel files.

Value sets for EQ-5D-3L are available for many countries and have been produced using the time trade-off (TTO) valuation technique or the visual analogue scale (VAS) valuation technique.  Some countries have TTO and VAS value sets for EQ-5D-3L.

For EQ-5D-5L, a standardised valuation study protocol (EQ-VT) was developed by the EuroQol group based on the composite time trade-off (cTTO) valuation technique supplemented by a discrete choice experiment (DCE).  The EuroQol group recommends users to use a standard value set where available.

The EQ-5D-5L "crosswalk" value sets published by [van Hout *et al*](https://pubmed.ncbi.nlm.nih.gov/22867780/) are also included. The crosswalk value sets enable index values to be calculated for EQ-5D-5L data where no value set is available by mapping between the EQ-5D-5L and EQ-5D-3L descriptive systems.

Additional information on EQ-5D can be found on the [EuroQol](https://euroqol.org/) website as well as in [Szende *et al* (2007)](https://www.doi.org/10.1007/1-4020-5511-0) and [Szende 
    *et al* (2014)](https://www.doi.org/10.1007/978-94-007-7596-1). Advice on [choosing a value set](https://euroqol.org/eq-5d-instruments/eq-5d-3l-about/valuation/choosing-a-value-set/
) can also be found on the EuroQol website.

## Installation

You can install the released version of eq5d from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("eq5d")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("fragla/eq5d")
```

## Quick Start

```{r quickStart}
library(eq5d)

#single calculation

#named vector MO, SC, UA, PD and AD represent mobility, self-care, usual activites, pain/discomfort and anxiety/depression, respectfully.
scores <- c(MO=1,SC=2,UA=3,PD=2,AD=1)

#EQ-5D-3L using the UK TTO value set
eq5d(scores=scores, country="UK", version="3L", type="TTO")

#Using five digit format
eq5d(scores=12321, country="UK", version="3L", type="TTO")

#multiple calculations using the Canadian VT value set

#data.frame with individual dimensions
scores.df <- data.frame(
  MO=c(1,2,3,4,5), SC=c(1,5,4,3,2), UA=c(1,5,2,3,1), PD=c(1,3,4,3,4), AD=c(1,2,1,2,1)
)

eq5d(scores.df, country="Canada", version="5L", type="VT")

#data.frame using five digit format
scores.df2 <- data.frame(state=c(11111,25532,34241,43332,52141))

eq5d(scores.df2, country="Canada", version="5L", type="VT", five.digit="state")

#or using a vector

eq5d(scores.df2$state, country="Canada", version="5L", type="VT")

```

## Value sets

The available value sets can be viewed using the ***valuesets*** function. The results can be filtered by EQ-5D version, value set type or by country.

```{r valuesets}
# Return all value sets (top 6 returned for brevity).
head(valuesets())

# Return VAS based value sets (top 6 returned for brevity).
head(valuesets(type="VAS"))

# Return EQ-5D-5L value sets (top 6 returned for brevity).
head(valuesets(version="5L"))

# Return all UK value sets.
valuesets(country="UK")
```

## Analysis of EQ-5D Profiles

A number of methods have been published that enable the analysis of EQ-5D profiles, most recently in the open access book Methods for Analysing and Reporting EQ-5D Data by [Devlin, Janssen and Parkin](https://www.springer.com/gp/book/9783030476212). The eq5d package includes R implentations of some of the methods from this book and from other sources that may be of use in analysing EQ-5D data.

### Cumulative frequency analysis

The ***eq5dcf*** function calculates the frequency, percentage, cumulative frequency and cumulative percentage for each five digit profile in an EQ-5D dataset. Either a vector of five digit profiles or a data.frame of indiviual dimensions can be passed to this function in order to summarise data in this way.

```{r eq5dcf}
library(readxl)

#load example data
data <- read_excel(system.file("extdata", "eq5d3l_example.xlsx", package="eq5d"))

#run eq5dcf function on a data.frame
res <- eq5dcf(data, "3L")

# Return data.frame of cumulative frequency stats (top 6 returned for brevity).
head(res)

```

### Summarising the Severity of EQ-5D Profiles

The eq5d package includes methods for summarising the severity of EQ-5D health state. The Level Sum Score (LSS) treats each dimension's level as a number rather than a category. Each number is added up to produce a score between 5 and 15 for EQ-5D-3L and 5 and 25 for EQ-5D-5L.

```{r eq5dlss}
lss(c(MO=1,SC=2,UA=3,PD=2,AD=1), version="3L")

lss(55555, version="5L")

lss(c(11111,12345, 55555), version="5L")

```

The Level Frequency Score (LFS) is an alternative method of summarising profile data developed by [Oppe and de Charro](https://pubmed.ncbi.nlm.nih.gov/11189116). Here the frequency of the levels for each health state are characterised. As described in Devlin, Janssen and Parkin's book, the full health profile 11111 for EQ-5D-5L has 5, 1 s, no level 2, 3, 4 and 5s, so the LFS is 50000; the health profile 55555 is 00005; profiles such as 31524 and 53412 would be 11111.

```{r eq5dlfs}
lfs(c(MO=1,SC=2,UA=3,PD=2,AD=1), version="3L")

lfs(55555, version="5L")

lfs(c(11111,12345, 55555), version="5L")

```

### Paretian Classification of Health Change

The Paretian Classification of Health Change (PCHC) was developed by [Devlin et al](https://pubmed.ncbi.nlm.nih.gov/20623685) in 2010 and is used to compare changes in individuals over time. PCHC classifies the change in an individual's health state as better (improvement in at least one dimension), worse (a deterioration in at least one dimension), mixed (improvements and deteriorations in dimensions) or there being no change in the health state. Those classified in the No change group with the 11111 health state can be separated into their own "No problems" group. PCHC can be calculated using the ***pchc*** function.

```{r pchc}
library(readxl)

#load example data
data <- read_excel(system.file("extdata", "eq5d3l_example.xlsx", package="eq5d"))

#use first 50 entries of each group as pre/post
pre <- data[data$Group=="Group1",][1:50,]
post <- data[data$Group=="Group2",][1:50,]

#run eq5dcf function on a data.frames

#Show no change, improve, worse, mixed without totals
res1 <- pchc(pre, post, version="3L", no.problems=FALSE, totals=FALSE)
res1

#Show totals, but not those with no problems
res2 <- pchc(pre, post, version="3L", no.problems=FALSE, totals=TRUE)
res2

#Show totals and no problems for each dimension
res3 <- pchc(pre, post, version="3L", no.problems=TRUE, totals=TRUE, by.dimension=TRUE)
res3

```

### EQ-5D-DS

The ***eq5dds*** function is an R approximation of the Stata command written by [Ramos-Goñi & Ramallo-Fariña](https://www.stata-journal.com/article.html?article=st0450). The function analyses and summarises the descriptive components of an EQ-5D dataset. The "by" argument enables a grouping variable to be specified when analysing the data subgroup.

```{r eq5dds}

dat <- data.frame(
         matrix(
           sample(1:3,5*12, replace=TRUE),12,5, 
           dimnames=list(1:12,c("MO","SC","UA","PD","AD"))
         ),
         Sex=rep(c("Male", "Female"))
       )

eq5dds(dat, version="3L")

eq5dds(dat, version="3L", counts=TRUE)

eq5dds(dat, version="3L", by="Sex")

```

## Helper functions

Helper functions are included, which may be useful in the processing of EQ-5D data. ***getHealthStates*** returns a vector of all possible five digit health states for a specified EQ-5D version. ***getDimensionsFromHealthStates*** splits a vector of five digit health states into a data.frame of their individual components and ***getHealthStatesFromDimensions*** combines indiviual dimensions in a data.frame into five digit health states..

```{r helper}

# Get all EQ-5D-3L five digit health states (top 6 returned for brevity).
head(getHealthStates("3L"))

# Split five digit health states into their individual components.
getDimensionsFromHealthStates(c("12345", "54321"), version="5L")

```

## Example data

Example data is included with the package and can be accessed using the ***system.file*** function.

```{r example data}
# View example files.
dir(path=system.file("extdata", package="eq5d"))

# Read example EQ-5D-3L data.
library(readxl)
data <- read_excel(system.file("extdata", "eq5d3l_example.xlsx", package="eq5d"))

# Calculate index scores
scores <- eq5d(data, country="UK", version="3L", type="TTO")

# Top 6 scores
head(scores)
```

## Shiny web interface

The calculation (and visualisation) of multiple EQ-5D indices can also be performed by upload of a CSV or Excel file using the packaged [Shiny](https://shiny.rstudio.com) app. This requires the [shiny](https://cran.r-project.org/package=shiny),  [DT](https://cran.r-project.org/package=DT), 
[FSA](https://cran.r-project.org/package=FSA), 
[ggplot2](https://cran.r-project.org/package=ggplot2), 
[ggiraph](https://cran.r-project.org/package=ggiraph), 
[ggiraphExtra](https://cran.r-project.org/package=ggiraphExtra), 
[mime](https://cran.r-project.org/package=mime),  [PMCMRplus](https://cran.r-project.org/package=PMCMRplus), [readxl](https://cran.r-project.org/package=readxl), 
[shinycssloaders](https://cran.r-project.org/package=shinycssloaders) and
[shinyWidgets](https://cran.r-project.org/package=shinyWidgets) packages. Ideally the CSV/Excel headers should be the same as the names of the vector passed to the ***eq5d*** function i.e. MO, SC, UA, PD and AD or the column name "State" if using the five digit format. However, a modal dialog will prompt the user to select the appropriate columns if the defaults can not be found.

The app is launched using the ***shiny_eq5d*** function.

```{r shiny, echo=TRUE, eval=FALSE}
shiny_eq5d()
```
Alternatively, it can be accessed without installing R/Shiny/eq5d by visiting [shinyapps.io](https://fragla.shinyapps.io/shiny-eq5d).

## License

This project is licensed under the MIT License - see the [LICENSE.md](https://github.com/fragla/eq5d/blob/master/LICENSE.md) file for details.

