---
title: "Reporting EQ-5D Severity and Distributional Summaries"
author: "Fraser Morton"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reporting EQ-5D Severity and Distributional Summaries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

In addition to descriptive system tables, EQ-5D data can be summarised using 
measures that describe the severity and distribution of observed health states.

This vignette introduces the severity and distributional reporting methods 
available in `eq5d`, including the Level Sum Score (LSS), Level Frequency 
Score (LFS), Shannon's entropy and evenness, and measures derived from health 
state cumulative frequency distributions.

## Example data

```{r}
suppressPackageStartupMessages(library(eq5d))

## Example EQ-5D-3L data included with the package
dat <- read.csv(
  system.file("extdata", "eq5d3l_example.csv", package = "eq5d")
)

## For an ungrouped population-level example
dat1 <- subset(dat, Group == "Group1")

## Example EQ-5D-3L profile (single observation)
scores <- unlist(dat1[1, c("MO", "SC", "UA", "PD", "AD")])
```

## Severity measures

Severity measures provide concise summaries of EQ-5D health states and can be 
reported alongside descriptive system tables.

The `eq5d` package provides two severity measures:

- the Level Sum Score (LSS)
- the Level Frequency Score (LFS)

```{r, eval=FALSE}
# Example severity summaries
lss(scores, version = "3L")
lfs(scores, version = "3L")
```

LSS is calculated by summing the reported levels across dimensions, with 
higher values indicating greater severity.

LFS provides a compact description of the EQ-5D profile based on the frequency 
of observed response levels within the health state.

These measures can be reported alongside descriptive system tables to provide 
additional information about health state severity.

## Visualising severity summaries

Severity measures such as LSS and LFS can also be used to summarise utility 
index values across severity strata. The plot_severity_summary() function 
provides a graphical representation of these relationships.

```{r eval=TRUE, fig.width = 6, fig.height = 6, fig.align = "center"}
plot_severity_summary(
  get_all_health_states("5L"),
  country = "UK",
  version = "5L",
  type = "VT",
  severity = "LSS"
)
```

Visual summaries can complement numerical severity measures by illustrating how 
utility values vary across levels of health-state severity. The plot below 
shows EQ-5D-5L utility values summarised by LSS. For each LSS value the lowest, 
median, and highest observed utility values are displayed.

## Shannon's entropy and evenness

The distribution of observed EQ-5D health states can also be summarised using 
Shannon's entropy and Shannon's evenness. These measures are calculated at the 
population level and therefore require a dataset of observed EQ-5D profiles.

```{r}
# Shannon's entropy and evenness for a population
shannon(dat1, version = "3L")
```

Shannon's entropy was first applied to EQ-5D data by Janssen et al. (2007)
as a way of describing the information contained within observed health state 
distributions. Shannon's entropy (H') reflects the diversity of observed 
responses while Shannon's evenness (J') measures how evenly responses are 
distributed relative to the maximum possible entropy.

These measures are often used to assess how much variation is 
present in the observed responses and can help identify concentration of 
responses arising from effects such as limited variation or ceiling effects.

## Health state distribution summaries

In addition to severity measures and entropy statistics, `eq5d` provides 
summaries of how health states are distributed within a population. These 
summaries are based on health state cumulative frequency data generated using 
`eq5dcf()`.

```{r}
# Health state cumulative frequency object
cf <- eq5dcf(dat1, version = "3L")
```

The **Health State Density Index (HSDI)** summarises the concentration of 
observed health states within a population. It can be useful when comparing the 
distribution of health states across different groups or datasets.

```{r}
idx <- hsdi(dat1, version = "3L")
idx
```

The Health State Density Curve (HSDC) provides a graphical representation of 
the cumulative distribution of health states.

```{r eval=TRUE, fig.width = 6, fig.height = 6, fig.align = "center"}
plot_hsdc(cf, hsdi = idx)
```

When interpreted alongside HSDI, the HSDC can help compare the distribution of 
health states across populations and identify differences in concentration 
across the health state spectrum.

## Comparing groups

When severity or distributional summaries are required for multiple groups, the 
data can be split before analysis. The helper functions below provide a 
convenient way to generate grouped HSDI and HSDC summaries.

```{r eval=TRUE, fig.width = 6, fig.height = 6, fig.align = "center"}
# Grouped HSDI
hsdi_by_group <- make_hsdi_by_group(
  dat,
  group = "Group",
  version = "3L"
)

hsdi_by_group

# Grouped HSDC data
hsdc_by_group <- make_hsdc_by_group(
  dat,
  group = "Group",
  version = "3L"
)

plot_hsdc(hsdc_by_group, hsdi = hsdi_by_group, group = "Group")
```

Grouped summaries can be useful when comparing the distribution of health 
states across study groups, populations or time points. These summaries are 
intended to complement descriptive system tables and other EQ-5D reporting 
outputs.

## Summary

This vignette introduced severity and distributional reporting methods
available in `eq5d`, including LSS, LFS, Shannon's entropy and evenness,
and health state distribution summaries based on cumulative frequency
distributions. These measures can complement descriptive system tables by
providing additional information about the severity, diversity and distribution 
of observed health states.
